Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-04 intro.tex
adding a citation
rephrased copied sentences

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{KS-entropy}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a periodic
signal, which exhibits low entropy, is on the low end of the
complexity spectrum.  We argue that an extension of \emph{permutation
  entropy}~\cite{bandt2002per}---a method for approximating the
entropy through ordinal analysis---is an effective way to assess the
complexity of a given time series.  Permutation entropy is ideal for
our purposes because it works with real-valued data and is known to
converge to the true entropy value. Other existing techniques either
require specific knowledge of the generating process or produce biased
values of the entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Rather than
use synthetic data, we study sensor data from computer performance
experiments.  While this is not a common laboratory experiment, it is
a highly appropriate choice here. {\color{red} [[Liz, these are the few sentences I was worried about being too similar but we may have only used them in a grant proposal.]]} Computers are among the most
complex engineered artifacts in current use.  Modern microprocessor
chips contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
loop that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.  
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a periodic
signal, which exhibits low entropy, is on the low end of the
complexity spectrum.  We argue that an extension of \emph{permutation
  entropy}~\cite{bandt2002per}---a method for approximating the
entropy through ordinal analysis---is an effective way to assess the
complexity of a given time series.  Permutation entropy is ideal for
our purposes because it works with real-valued data and is known to
converge to the true entropy value. Other existing techniques either
require specific knowledge of the generating process or produce biased
values of the entropy~\cite{bollt2001}.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Rather than
use synthetic data, we study sensor data from computer performance
experiments.  While this is not a common laboratory experiment, it is
a highly appropriate choice here.  Computers are extremely complicated
systems.  Modern microprocessor chips contain multiple processing
units and multi-layer memories, for instance, and they use complicated
hardware/software strategies to maximize performance by moving data
and threads of computation across those resources.  As a result, the
processor and memory loads during the execution of even a very simple
program can evolve in a very complicated fashion.
Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
loop that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh
